**Supplementary Information**

### 2020-09-15 submission/manuscript.Rmd
Coauthor review #2.

An alternative to these procedures is a direct isolation of the targeted material by incubating macrophyte tissues in an extraction buffer. After the incubation, the undisrupted host tissue is removed and the isolation procedure continues, omitting host material contaminations. To our knowledge the only procedure describing a direct and selective epiphytic DNA isolation from the surfaces of marine macrophytes was provided by @Burke2009. In contrast to previously described methods this protocol enables an almost complete removal of the surface community and it was used for 16S rRNA gene clone library construction [@Burke2011a] and metagenome sequencing [@Burke2011]. This method, thought providing a selective isolation procedure, uses a rapid multienzyme cleaner (3M) that is not available worldwide and without a known composition [@Burke2009]. Also to our knowledge, no selective isolation protocol for proteins from epiphytic communities inhabiting marine macrophytes was established.

@@

Obtained sequences were analyzed on the computer cluster Isabella (University Computing Center, University of Zagreb) using mothur (version 1.43.0) [@Schloss2009] according to the MiSeq Standard Operating Procedure (MiSeq SOP; https://mothur.org/wiki/MiSeq_SOP) [@Kozich2013] and recommendations given from the Riffomonas project to enhance data reproducibility (http://www.riffomonas.org/). For alignment and classification of sequences the SILVA SSU Ref NR 99 database (release 138; http://www.arb-silva.de) was used [@Quast2013; @Yilmaz2014]. Sequences classified as chloroplasts by SILVA were exported and reclassifed using mothur and the RDP (http://rdp.cme.msu.edu/) training set (version 16) reference files adapted for mothur [@Cole2014]. In comparison to SILVA, RDP allows a more detailed classification of chloroplast sequences. Pipeline data processing and visualization was done using R (version 3.6.0) [@RCoreTeam2019],  package tidyverse (version 1.3.0) [@Wickham2019] and multiple other packages [@Xie2014; @Xie2015; @Xie2019a; @Xie2019; @Xie2020; @Allaire2019; @Xie2018; @Zhu2019; @Neuwirth2014; @Bengtsson2020; @Wilke2018]. The detailed analysis procedure including the R Markdown file for this paper are available as a GitHub repository (https://github.com/mkorlevic/Korlevic_SelectiveRemoval_EnvironMicrobiol_2020). Based on the ATCC MSA-1002 mock community included in the analysis a sequencing error rate of `r error` \si{\percent} was determined, which is in line with previously reported values for next-generation sequencing data [@Kozich2013; @Schloss2016]. In addition, the negative control processed together with the samples yielded only `r nc` sequences after sequence quality curation.

@@

@@

>>>>

An alternative to these procedures is a direct isolation of the targeted material by incubating macrophyte tissues in an extraction buffer. After the incubation, the undisrupted host tissue is removed and the isolation procedure continues, omitting host material contaminations. To our knowledge the only procedure describing a direct and selective epiphytic DNA isolation from the surfaces of marine macrophytes was provided by @Burke2009. In contrast to previously described methods this protocol enables an almost complete removal of the surface community and it was used for 16S rRNA gene clone library construction [@Burke2011a] and metagenome sequencing [@Burke2011]. This method, though providing a selective isolation procedure, uses a rapid multienzyme cleaner (3M) that is not available worldwide and without a known composition [@Burke2009]. Also to our knowledge, no selective isolation protocol for proteins from epiphytic communities inhabiting marine macrophytes was established.

@@

Obtained sequences were analyzed on the computer cluster Isabella (University Computing Center, University of Zagreb) using mothur (version 1.43.0) [@Schloss2009] according to the MiSeq Standard Operating Procedure (MiSeq SOP; https://mothur.org/wiki/MiSeq_SOP) [@Kozich2013] and recommendations given from the Riffomonas project to enhance data reproducibility (http://www.riffomonas.org/). For alignment and classification of sequences the SILVA SSU Ref NR 99 database (release 138; http://www.arb-silva.de) was used [@Quast2013; @Yilmaz2014]. Sequences classified as chloroplasts by SILVA were exported and reclassified using mothur and the RDP (http://rdp.cme.msu.edu/) training set (version 16) reference files adapted for mothur [@Cole2014]. In comparison to SILVA, RDP allows a more detailed classification of chloroplast sequences. Based on the ATCC MSA-1002 mock community included in the analysis a sequencing error rate of `r error` \si{\percent} was determined, which is in line with previously reported values for next-generation sequencing data [@Kozich2013; @Schloss2016]. In addition, the negative control processed together with the samples yielded only `r nc` sequences after sequence quality curation.

@@

### Data processing and visualization
Processing and visualization  of 16S rRNA, metagenomic and metaproteomic data was done using R (version 3.6.0) [@RCoreTeam2019],  package tidyverse (version 1.3.0) [@Wickham2019] and multiple other packages [@Xie2014; @Xie2015; @Xie2019a; @Xie2020; @Allaire2019; @Xie2018; @Zhu2019; @Neuwirth2014; @Bengtsson2020; @Wilke2018]. The detailed analysis procedure including the R Markdown file for this paper are available as a GitHub repository (https://github.com/mkorlevic/Korlevic_SelectiveRemoval_EnvironMicrobiol_2020).

@@

**\autoref{cog}.** \nameref{cog}

### 2020-09-15 submission/manuscript.tex
Coauthor review #2.

An alternative to these procedures is a direct isolation of the targeted
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed and the isolation
procedure continues, omitting host material contaminations. To our
knowledge the only procedure describing a direct and selective epiphytic
DNA isolation from the surfaces of marine macrophytes was provided by
Burke \emph{et al.} (2009). In contrast to previously described methods
this protocol enables an almost complete removal of the surface
community and it was used for 16S rRNA gene clone library construction
(Burke and Thomas \emph{et al.}, 2011) and metagenome sequencing (Burke
and Peter Steinberg \emph{et al.}, 2011). This method, thought providing
a selective isolation procedure, uses a rapid multienzyme cleaner (3M)
that is not available worldwide and without a known composition (Burke
\emph{et al.}, 2009). Also to our knowledge, no selective isolation
protocol for proteins from epiphytic communities inhabiting marine
macrophytes was established.

@@

Obtained sequences were analyzed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss \emph{et al.}, 2009) according to the MiSeq
Standard Operating Procedure (MiSeq SOP;
\url{https://mothur.org/wiki/MiSeq_SOP}) (Kozich \emph{et al.}, 2013)
and recommendations given from the Riffomonas project to enhance data
reproducibility (\url{http://www.riffomonas.org/}). For alignment and
classification of sequences the SILVA SSU Ref NR 99 database (release
138; \url{http://www.arb-silva.de}) was used (Quast \emph{et al.}, 2013;
Yilmaz \emph{et al.}, 2014). Sequences classified as chloroplasts by
SILVA were exported and reclassifed using mothur and the RDP
(\url{http://rdp.cme.msu.edu/}) training set (version 16) reference
files adapted for mothur (Cole \emph{et al.}, 2014). In comparison to
SILVA, RDP allows a more detailed classification of chloroplast
sequences. Pipeline data processing and visualization was done using R
(version 3.6.0) (R Core Team, 2019), package tidyverse (version 1.3.0)
(Wickham \emph{et al.}, 2019) and multiple other packages (Xie, 2014,
2015, 2020; Neuwirth, 2014; Xie \emph{et al.}, 2018; Wilke, 2018; Y.
Xie, 2019b, 2019a; Allaire \emph{et al.}, 2019; Zhu, 2019; Bengtsson,
2020). The detailed analysis procedure including the R Markdown file for
this paper are available as a GitHub repository
(\url{https://github.com/mkorlevic/Korlevic_SelectiveRemoval_EnvironMicrobiol_2020}).
Based on the ATCC MSA-1002 mock community included in the analysis a
sequencing error rate of 0.009 \si{\percent} was determined, which is in
line with previously reported values for next-generation sequencing data
(Kozich \emph{et al.}, 2013; Schloss \emph{et al.}, 2016). In addition,
the negative control processed together with the samples yielded only 2
sequences after sequence quality curation.

@@

@@

\leavevmode\hypertarget{ref-Xie2019}{}%
Xie, Y. (2019a) knitr: A general-purpose package for dynamic report
generation in R.

\leavevmode\hypertarget{ref-Xie2019a}{}%
Xie, Y. (2019b) TinyTeX: A lightweight, cross-platform, and
easy-to-maintain LaTeX distribution based on TeX Live. \emph{TUGboat}
\textbf{40}: 30--32.

@@

>>>>

An alternative to these procedures is a direct isolation of the targeted
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed and the isolation
procedure continues, omitting host material contaminations. To our
knowledge the only procedure describing a direct and selective epiphytic
DNA isolation from the surfaces of marine macrophytes was provided by
Burke \emph{et al.} (2009). In contrast to previously described methods
this protocol enables an almost complete removal of the surface
community and it was used for 16S rRNA gene clone library construction
(Burke and Thomas \emph{et al.}, 2011) and metagenome sequencing (Burke
and Peter Steinberg \emph{et al.}, 2011). This method, though providing
a selective isolation procedure, uses a rapid multienzyme cleaner (3M)
that is not available worldwide and without a known composition (Burke
\emph{et al.}, 2009). Also to our knowledge, no selective isolation
protocol for proteins from epiphytic communities inhabiting marine
macrophytes was established.

@@

Obtained sequences were analyzed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss \emph{et al.}, 2009) according to the MiSeq
Standard Operating Procedure (MiSeq SOP;
\url{https://mothur.org/wiki/MiSeq_SOP}) (Kozich \emph{et al.}, 2013)
and recommendations given from the Riffomonas project to enhance data
reproducibility (\url{http://www.riffomonas.org/}). For alignment and
classification of sequences the SILVA SSU Ref NR 99 database (release
138; \url{http://www.arb-silva.de}) was used (Quast \emph{et al.}, 2013;
Yilmaz \emph{et al.}, 2014). Sequences classified as chloroplasts by
SILVA were exported and reclassified using mothur and the RDP
(\url{http://rdp.cme.msu.edu/}) training set (version 16) reference
files adapted for mothur (Cole \emph{et al.}, 2014). In comparison to
SILVA, RDP allows a more detailed classification of chloroplast
sequences. Based on the ATCC MSA-1002 mock community included in the
analysis a sequencing error rate of 0.009 \si{\percent} was determined,
which is in line with previously reported values for next-generation
sequencing data (Kozich \emph{et al.}, 2013; Schloss \emph{et al.},
2016). In addition, the negative control processed together with the
samples yielded only 2 sequences after sequence quality curation.

@@

\hypertarget{data-processing-and-visualization}{%
\subsubsection{Data processing and
visualization}\label{data-processing-and-visualization}}

Processing and visualization of 16S rRNA, metagenomic and metaproteomic
data was done using R (version 3.6.0) (R Core Team, 2019), package
tidyverse (version 1.3.0) (Wickham \emph{et al.}, 2019) and multiple
other packages (Xie, 2014, 2015, 2019, 2020; Neuwirth, 2014; Xie
\emph{et al.}, 2018; Wilke, 2018; Allaire \emph{et al.}, 2019; Zhu,
2019; Bengtsson, 2020). The detailed analysis procedure including the R
Markdown file for this paper are available as a GitHub repository
(\url{https://github.com/mkorlevic/Korlevic_SelectiveRemoval_EnvironMicrobiol_2020}).

@@

\leavevmode\hypertarget{ref-Xie2019a}{}%
Xie, Y. (2019) TinyTeX: A lightweight, cross-platform, and
easy-to-maintain LaTeX distribution based on TeX Live. \emph{TUGboat}
\textbf{40}: 30--32.

@@

\textbf{\autoref{cog}.} \nameref{cog}

### 2020-09-15 submission/supplementary.Rmd
Coauthor review #2.

# Generating the table
select(metadata_metagenomic_phylum, phylum, cds, cds_relative, rpkm, rpkm_relative) %>%
kable(format="latex", booktabs=TRUE, linesep="", longtable=TRUE, escape=FALSE, col.names=linebreak(c("Phylum", "No. of CDS", "CDS (\\%)", "Summed RPKM", "RPKM (\\%)"), align="c"), caption="Phyla into which CDS were classifed, number and proportion of CDS in different phyla and sum of coding sequences' RPKM (Reads Per Kilobase Million) and their proportion in different phyla. Data are derived from sequenced metagenomes. Each metagenomic sample is labeled with sampling station, community type, sampling date and sample ID. For each sample top ten phyla based on RPKM were selected. CDS that were not successfully classified were excluded from the dataset.\\label{metagenomic_taxonomy}", align="c") %>%
  kable_styling(latex_options=c("hold_position", "repeat_header"), font_size=9, full_width=TRUE) %>%
  row_spec(0, bold=TRUE) %>%
  group_rows("Funtana, \\\\textit{Cymodocea nodosa} (Monospecific), 14 December 2017 (sample ID: 45)", 1, 10, escape=FALSE, latex_gap_space="0.6 em") %>%
  group_rows("Funtana, \\\\textit{Caulerpa cylindracea} (Monospecific), 14 December 2017 (sample ID: 47)", 11, 20, escape=FALSE, latex_gap_space="0.6 em") %>%
  group_rows("Funtana, \\\\textit{Cymodocea nodosa} (Mixed), 19 June 2018 (sample ID: 61)", 21, 30, escape=FALSE, latex_gap_space="0.9 em") %>%
  group_rows("Funtana, \\\\textit{Caulerpa cylindracea} (Monospecific), 19 June 2018 (sample ID: 63)", 31, 40, escape=FALSE, latex_gap_space="0.6 em")
```

>>>>

# Generating the table
select(metadata_metagenomic_phylum, phylum, cds, cds_relative, rpkm, rpkm_relative) %>%
kable(format="latex", booktabs=TRUE, linesep="", longtable=TRUE, escape=FALSE, col.names=linebreak(c("Phylum", "No. of CDS", "CDS (\\%)", "Summed RPKM", "RPKM (\\%)"), align="c"), caption="Phyla into which CDS were classifed, number and proportion of CDS in different phyla and sum of coding sequences' RPKM (Reads Per Kilobase Million) and their proportion in different phyla. Data are derived from sequenced metagenomes. Each metagenomic sample is labeled with sampling station, community type, sampling date and sample ID. For each sample top ten phyla based on RPKM were selected. CDS that were not successfully classified were excluded from the dataset.\\label{metagenomic_taxonomy}", align="c") %>%
  kable_styling(latex_options=c("hold_position", "repeat_header"), font_size=9, full_width=TRUE) %>%
  row_spec(0, bold=TRUE) %>%
  group_rows("Funtana, \\\\textit{Cymodocea nodosa} (Mixed), 14 December 2017 (sample ID: 45)", 1, 10, escape=FALSE, latex_gap_space="0.6 em") %>%
  group_rows("Funtana, \\\\textit{Caulerpa cylindracea} (Monospecific), 14 December 2017 (sample ID: 47)", 11, 20, escape=FALSE, latex_gap_space="0.6 em") %>%
  group_rows("Funtana, \\\\textit{Cymodocea nodosa} (Mixed), 19 June 2018 (sample ID: 61)", 21, 30, escape=FALSE, latex_gap_space="0.9 em") %>%
  group_rows("Funtana, \\\\textit{Caulerpa cylindracea} (Monospecific), 19 June 2018 (sample ID: 63)", 31, 40, escape=FALSE, latex_gap_space="0.6 em")
```
